---
title: "Combining all data"
author: "Sophia Harris, Sam McFarland, Esal Shakil, Nika Yermakova"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(usmap)
library(writexl)
library(sf)
library(readxl)
```

### combining all the obesity files together

Each state has it's own file with its counties.
Note: We modified the files for alaska and lousiana to match the column names
with the other files since they don't have "counties".

```{r, message=FALSE, results = FALSE, warning=FALSE}
# getting the list of file names
file_list <- list.files(pattern = "\\.csv$")

# extracting state names from file names and capitalizing only the first letter
state_names <- gsub("^(.*)-\\d{4}-.*$", "\\1", file_list)
state_names <- str_to_title(state_names)

# replacing dashes with spaces for states with two-word names
state_names <- gsub("-", " ", state_names)

# reading each CSV file into a list of data frames
state_data <- lapply(file_list, read_csv)

# creating a column for state names in each data frame
state_data_with_names <- Map(function(df, state) {
  df$State <- state
  df
}, state_data, state_names)

# combining all data frames into one
combined_df <- do.call(rbind, state_data_with_names)

# re-ording the position of the columns
combined_obesity <- combined_df |>
  select(State, County, `County Value`, `Error Margin`, `Z-Score`) |>
  rename(`Obesity Rate` = `County Value`)

# making another column where state and county are combined
combined_obesity$Combined <- paste(combined_obesity$County, combined_obesity$State, sep = "_")
```

### combining food insecurity data with county data to get state info

```{r, message=FALSE, results = FALSE, warning=FALSE}
# loading in food insecurity data from the urban institute
food_insecurity <- read_excel("chart_data.xlsx")

# loading in county data from us_map()
county_info <- us_map(regions = "county")

# combining food insecurity and county data
county_info <- county_info |>
  mutate("fips" = as.numeric(fips))

food_insecurity <- food_insecurity |>
  mutate("id" = as.numeric(id))

food_insecurity <- left_join(food_insecurity, county_info,
c("id"="fips"))
```

### loading in + tidying population data

We don't actually need this for our final data. However, we initially included
this because we thought that we would be doing something with state data and 
needed the population information to configure weighted averages. 

While this data is irrelevant, it was part of the process in getting the data
combined so we are keeping this piece. 


```{r, message=FALSE, results = FALSE, warning=FALSE}
# loading in population data
population_info <- read_excel("co-est2019-annres.xlsx")

# renaming the columns
population_info <- population_info[-(1:3), ]
    
# Renaming the first column to "select characteristics"
names(population_info)[1] <- "Geographic Area"
names(population_info)[2] <- "Census"
names(population_info)[3] <- "Estimates Base"
names(population_info)[4] <- "2010"
names(population_info)[5] <- "2011"
names(population_info)[6] <- "2012"
names(population_info)[7] <- "2013"
names(population_info)[8] <- "2014"
names(population_info)[9] <- "2015"
names(population_info)[10] <- "2016"
names(population_info)[11] <- "2017"
names(population_info)[12] <- "2018"
names(population_info)[13] <- "2019"

# Converting chr data into num
population_info <- population_info |>
  mutate(across(c("Census", "Estimates Base", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019"), as.numeric))

# Separate state from county in the name and combine with underscore
population_info <- population_info |>
  separate(col = "Geographic Area", 
           into = c("County", "State_temp"),
           sep = ",",
           remove = FALSE) |> 
  mutate(County = gsub("^\\.", "", County),  # Remove leading period
         State_temp = gsub("^\\s+|\\s+$", "", State_temp),  # Remove leading/trailing whitespace from state
         State = gsub(" ", "_", State_temp)) |>  # Replace spaces in state name with underscores
  select(-State_temp)  # Remove temporary state column

# Combine county and state names with an underscore
population_info <- population_info |>
  mutate(Unique_ID = paste0(County, "_", State))

# renaming the 2016 data to be our population of interest
population_info <- population_info |>
  rename(population = `2016`)
```

### joining food insecurity with population data

```{r, message=FALSE, results = FALSE, warning=FALSE}
# creating unique id for food insecurity data 
food_insecurity$Unique_ID <- paste(food_insecurity$name, food_insecurity$full, sep = "_")

# cleaning up labels so they match up better for merging
food_insecurity <- food_insecurity |>
  mutate(Unique_ID = gsub("County_NA", "County_Connecticut", Unique_ID)) |>
  mutate(Unique_ID = gsub("Census Area_NA", "Census Area_Alaska", Unique_ID))

population_info <- population_info|>
  mutate(Unique_ID = gsub(" city_", " City_", Unique_ID)) |>
  mutate(Unique_ID = gsub("District_of_Columbia", "District of Columbia", Unique_ID)) |>
  mutate(Unique_ID = gsub("New_Hampshire", "New Hampshire", Unique_ID))  |>
  mutate(Unique_ID = gsub("New_Jersey", "New Jersey", Unique_ID))  |>
  mutate(Unique_ID = gsub("New_Mexico", "New Mexico", Unique_ID))  |>
  mutate(Unique_ID = gsub("New_York", "New York", Unique_ID)) |>
  mutate(Unique_ID = gsub("_Carolina", " Carolina", Unique_ID)) |>
  mutate(Unique_ID = gsub("_Dakota", " Dakota", Unique_ID)) |>
  mutate(Unique_ID = gsub("Rhode_Island", "Rhode Island", Unique_ID)) |>
  mutate(Unique_ID = gsub("West_Virginia", "West Virginia", Unique_ID))

# merging data
food_insecurity <- left_join(food_insecurity, population_info, by = "Unique_ID")
```

### combining food insecurity with obesity data

```{r, message=FALSE, results = FALSE, warning=FALSE}
# mutating some of the names
food_insecurity <- food_insecurity |>
  mutate(Unique_ID = gsub(" County", "", Unique_ID)) |>
  mutate(Unique_ID = gsub(" Borough", "", Unique_ID)) |>
  mutate(Unique_ID = gsub(" Census Area", "", Unique_ID)) |>
  mutate(Unique_ID = gsub(" Parish", "", Unique_ID)) |>
  mutate(Unique_ID = gsub(" Municipality", "", Unique_ID)) |>
  mutate(Unique_ID = gsub(" City and", "", Unique_ID))

combined_obesity <- combined_obesity |>
  mutate(Combined = gsub("^_", "_", Combined))

# left joining the data sets based on the formatted columns
final_data <- left_join(combined_obesity, food_insecurity, by = c("Combined" = "Unique_ID")) |>
  filter(!is.na(`Geographic Area`)) |> 
  mutate(`Obesity Rate` = as.numeric(gsub("%", "", `Obesity Rate`))) |>
  rename(`Food Insecurity` = "food_insecure_all")
  

# tidying final data
final_data <- final_data |>
  select(State.x, abbr, name, `Geographic Area`, `Obesity Rate`, `Food Insecurity`, food_insecure_children, low_birthweight, diabetes, disability,
         no_insurance, housing_cost_burdened, severely_housing_cost_burdened, wage_fair_market_rent, median_income, below_poverty,
         unemployment, credit_score, debt, children, seniors, people_color, college_less, rural_population, geom) |>
  rename(state = State.x,
         county = name,
         obesity_rate = `Obesity Rate`,
         food_insecurity_rate = `Food Insecurity`) |>
  mutate(across(c(food_insecure_children, low_birthweight, diabetes, disability, no_insurance, housing_cost_burdened,
                  severely_housing_cost_burdened, below_poverty, unemployment, debt, children, seniors,
                  people_color, college_less, rural_population, credit_score, wage_fair_market_rent), as.numeric))|>
  mutate(`Geographic Area` = sub("^\\.", "", `Geographic Area`)) |>
  rename(geographic_area = `Geographic Area`) |>
  select(-geom)
```

### turning data frame into excel file

```{r, message=FALSE, results = FALSE, warning=FALSE}
# exporting the data to excel
write_xlsx(final_data, "final_data.xlsx")
```

